---
title: "Final Project: Baseball Over the Last 40 Years"
author: "Samuel Harris"
format: html
editor: visual
theme: simple
---

## Research Question + Variables

-   **Research Question:** How has baseball changed in the past 40 years, and why has it changed?

-   I think this is interesting because baseball has taken a more statistical turn in the past 20 years, and I believe this should be reflected in the data. Teams started to value players based on different metrics, and I believe those desired metrics have increased over time in overall team stats.

-   The data I'm using is from **Lahman's Baseball Database** which is found here, http://seanlahman.com/download-baseball-database/. I am focusing on a subset from Lahman's Database called "teams" which contains various team statistics like home runs and strikeouts each year from 1871 to 2022.

-   Here's a glance at the summary statistics of the numeric variables I will be focusing on from the data set. I chose these variables since they are fundamental in baseball. **The data has been cleaned to only contain team statistics from 1980 onward, and statistics are normalized to per game.**

```{r, echo=FALSE, message=FALSE}
library(gridExtra)
library(corrplot)
library(tidyverse)
library(estimatr)
library(skimr)
library(tidyr)

rm(list=ls())
b = read_csv("Teams.csv", show_col_types = FALSE)

#Add win percentage
b = b %>% 
  mutate(winpct = W/G)

#removing IPouts
b = b %>% 
  select(-IPouts)

# normalizing variables to per game and subseting from 1980 on
b <- b %>%
  mutate_at(vars(15:28,30:37), ~ round(. / b$G, 2))

b = b[b$yearID >= 1980,]
b = b %>% 
   select(yearID, name, franchID, H, HR, SO, R, winpct)

b %>% 
  skim() %>% 
  yank('numeric')
```

### Variables over Time

-   To look at which variables have changed over time, I calculated the mean for each variable grouped by year. I put these means into a data frame so that league-wide means for each are displayed in the data frame.
-   The outcome variables of interest over time are hits, home runs, and strikeouts.

```{r, echo=FALSE}
# Create new data set with means by year

summary_stats <- b %>%
  group_by(yearID) %>%
  summarise_at(vars(3:5), mean)

knitr::kable(summary_stats, digits = 2, caption="League-wide Means by Year")
```

## Correlation Over Time

-   To visualize correlation between variables, I created a correlation plot for all variables.

```{r, echo=FALSE}
## Create Correlation Plot

corr_matrix <- cor(summary_stats)

corrplot(corr_matrix, type = "upper", order = "hclust", tl.col = "black",addCoef.col = "white", number.cex = 1, method='color')
```

-   I then created a table that showed the correlation from each variable to the year variable. **Strikeouts were highly correlated,** increasing steadily over the past 40 years.

```{r, echo=FALSE}
#### Create Dataframe
result <- data.frame(Variable=character(), Cor=numeric(), Pval=numeric(), stringsAsFactors=FALSE)

for (i in 2:4) {
  corr <- cor.test(summary_stats[[i]], summary_stats$yearID)
  result[i, "Variable"] <- colnames(summary_stats)[i]
  result[i, "Cor"] <- round(corr$estimate, 2)
  result[i, "Pval"] <- round(corr$p.value, 2)
}
result <- na.omit(result)

## Show Variables correlation over time
result = result %>% 
   arrange(desc(abs(Cor)))

knitr::kable(result, caption= "Correlation to Year Table")
```

### Scatter Plots

-   These scatter plots visualize each variable's correlation with time.

-   Again, it's easily seen that strikeouts have steadily increased over time. But **why have strikeouts increased?** Aren't strikeouts a reason why teams lose? Why would teams want to strikeout more? This is the question I will be focusing on next.

```{r, echo=FALSE, message=FALSE}
###Scatter plots 
df = summary_stats

# Convert the data frame to long format
df_long <- df %>%
  pivot_longer(-yearID, names_to = "variable", values_to = "value")

# Create a scatter plot for each variable
df_long %>%
  ggplot(aes(x = yearID, y = value)) + 
  geom_smooth(method = "lm")+
  geom_point() + 
  labs(x = "Time", y ="Statistic",
       caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
  ggtitle("Per Game Baseball Stats Over Time") +
  facet_wrap(~ variable, scales = "free_y") +
   theme_bw() +
  theme(plot.caption = element_text(size =8, hjust = 0, vjust= -2.1, face="italic")) +
  theme(plot.title = element_text(hjust = 0.5, vjust = 2))

```

## Strikeouts {scrollable=""}

-   New question: What variables predict strikeouts?

-   To see which variables predict strikeouts, I performed a regression analysis. A model containing yearID, hits, and home runs as predictors for strikeouts had a relatively high r squared value of .79.

-   This model also showed a strong positive relationship between home runs and strikeouts, and a weaker negative relationship between hits and strikeouts.

    ```{r, echo=FALSE, message=FALSE, warning=FALSE}
    library(teamcolors)
    library(modelsummary)
    library(Lahman)
    library(gganimate)
    library(gapminder)
    library(gifski)
    library(plotly)
    # What are stikeouts correlated with?

    b1 <- b


    # Cleaning Names
    b1 = b1 %>% 
    mutate(name = ifelse(name == "Los Angeles Angels of Anaheim", "Los Angeles Angels", name),
           name = ifelse(name == "California Angels", "Los Angeles Angels", name),
           name = ifelse(name == "Anaheim Angels", "Los Angeles Angels", name),
           name = ifelse(name == "Tampa Bay Devil Rays", "Tampa Bay Rays", name),
           name = ifelse(name == "Cleveland Guardians", "Cleveland Indians", name),
           name = ifelse(name == "Montreal Expos", "Washington Nationals", name),
           name = ifelse(name == "Florida Marlins", "Miami Marlins", name))


    modelsummary(lm(SO ~ yearID + H + HR, data = b1))
    ```

### Visualizations

-   These plots visualize the relationship between strikeouts and home runs and strikeouts and hits. The plots also have win percentage indicated by size, team indicated by color, and a slider for the year.
-   The positive relationship between home runs and strikeouts is clearly seen.
-   The negative relationship between hits and strikeouts is also clearly seen.

```{r, echo=FALSE, warning=FALSE}
# Recently, as strikeouts have increased hits have decreased yet homeruns increased.
# Lets visualize this. 

g1 = b1  %>% 
  ggplot(aes(x = SO, y = HR, color = name, fill = name)) +
  geom_point(aes(frame = yearID, size = winpct))+
   scale_fill_teams(guide = FALSE) +
  scale_color_teams(2, guide = FALSE) +
  labs(x = "Strikeouts", 
       y = "Homeruns", 
       title  = "Strikeouts vs Homeruns by Team since 1980 (Per Game)",
       caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
  theme_bw() +
    theme(legend.key.height= unit(1.2, 'cm'),
        legend.key.width= unit(1, 'cm'))

ggplotly(g1) %>% 
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = .4, y = -.24, text = "Source: Lahman’s Baseball Database - \"Teams\"", xref='paper', yref='paper', showarrow = F,  xanchor='right', yanchor='auto', xshift=0, yshift=0,font = list(size = 7)))

g2 = b1  %>% 
  ggplot(aes(x = SO, y = H, color = name, fill = name)) +
  geom_point(aes(frame = yearID, size = winpct))+
   scale_fill_teams(guide = FALSE) +
  scale_color_teams(2, guide = FALSE) +
  labs(x = "Strikeouts", 
       y = "Hits", 
       title  = "Strikeouts vs Hits by Team since 1980 (Per Game)",
       caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
  theme_bw() +
    theme(legend.key.height= unit(1.2, 'cm'),
        legend.key.width= unit(1, 'cm')) 

ggplotly(g2)%>% 
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = .4, y = -.24, text = "Source: Lahman’s Baseball Database - \"Teams\"", xref='paper', yref='paper', showarrow = F,  xanchor='right', yanchor='auto', xshift=0, yshift=0,font = list(size = 7)))
```

## Runs

-   So, what's more valuable for scoring runs, hits or home runs? A model containing hits, and home runs as predictors for runs had a relatively high r squared value of .82.

-   This model also showed a strong positive relationship between home runs and runs, and a weaker positive relationship between hits and runs. The next tab visualizes the relationships.

```{r, echo=FALSE, warning=FALSE}
modelsummary(lm(R ~HR + H, data = b1))
```

### Visualizations

-   The first plot shows a clear linear relationship between home runs and runs.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Looking at the relationship between Hits and Runs and Homeruns and Runs

g3 = b1  %>% 
  ggplot(aes(x = HR, y = R, fill = winpct)) +
  geom_point(shape =21, size =1.8)+
  labs(x = "Homeruns per Game", 
       y = "Runs per Game", 
       title  = "Homeruns vs Runs by Team since 1980",
       fill = 'Win Percentage',
       caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
  theme_bw() +
    theme(legend.key.height= unit(1.2, 'cm'),
        legend.key.width= unit(1, 'cm')) +
    theme(plot.caption = element_text(size =8, hjust = 0, vjust= -2.1, face="italic"))+ scale_fill_gradient(low = "yellow", high = "blue") +
  geom_smooth(color="red") 

g3
```

-   However, the second plot shows a different relationship between hits and runs. Runs stay constant as hits increases until hits reach a value of about 8.5. After 8.5 hits, the relationship seems more linear.

-   So, this analysis shows that a team must reach a certain amount of hits before they begin scoring more runs. It also shows that home runs always guarantee that more runs are scored.

    ```{r, echo= FALSE, warning=FALSE, message=FALSE}
    g4 = b1  %>% 
      ggplot(aes(x = H, y = R, fill = winpct)) +
      geom_point(shape =21, size =1.8)+
      labs(x = "Hits per Game", 
           y = "Runs per Game", 
           title  = "Hits vs Runs by Team since 1980",
           fill = 'Win Percentage',
           caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
      theme_bw() +
        theme(legend.key.height= unit(1.2, 'cm'),
            legend.key.width= unit(1, 'cm')) +
        theme(plot.caption = element_text(size =8, hjust = 0, vjust= -2.1, face="italic"))+ scale_fill_gradient(low = "yellow", high = "blue") +
        geom_smooth(color="red")

    g4
    ```

## Conclusions {scrollable=""}

-   So, the final conclusion is that **strikeouts have increased over time because teams have begun to value home runs more than hits.** Teams are willing to to strikeout more if it means they hit more home runs which translate to guaranteed runs.

-   These final plots visually show the increase of strikeouts and home runs over time. There is a slider to view a particular team.

    ```{r, echo=FALSE, warning=FALSE}
    #So, yes. Teams have realized the value of home runs is greater than general hits like singles.
    #So, over time, teams have built teams that have hit more home runs while also striking out more.

    g5 = b1  %>% 
      ggplot(aes(x = yearID, y = HR, color= name)) +
      geom_line(aes(frame = franchID))+
      labs(x = "Year", 
           y = "Homeruns per Game", 
           title  = "Homeruns by Team since 1980",
           caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
      scale_color_teams() +
      theme_bw() +
        theme(legend.key.height= unit(1.2, 'cm'),
            legend.key.width= unit(1, 'cm')) +
        theme(plot.caption = element_text(size =8, hjust = 0, vjust= -2.1, face="italic"))

    ggplotly(g5)%>% 
      layout(margin = list(l = 50, r = 50, b = 100, t = 50),
             annotations = list(x = .4, y = -.24, text = "Source: Lahman’s Baseball Database - \"Teams\"", xref='paper', yref='paper', showarrow = F,  xanchor='right', yanchor='auto', xshift=0, yshift=0,font = list(size = 7)))

    g6 = b1  %>% 
      ggplot(aes(x = yearID, y = SO, color= name)) +
      geom_line(aes(frame = franchID))+
      labs(x = "Year", 
           y = "Strikeouts per Game", 
           title  = "Strikeouts by Team since 1980",
           caption = "Source: Lahman’s Baseball Database - \"Teams\"") +
      scale_color_teams() +
      theme_bw() +
        theme(legend.key.height= unit(1.2, 'cm'),
            legend.key.width= unit(1, 'cm')) 

    ggplotly(g6)%>% 
      layout(margin = list(l = 50, r = 50, b = 100, t = 50),
             annotations = list(x = .4, y = -.24, text = "Source: Lahman’s Baseball Database - \"Teams\"", xref='paper', yref='paper', showarrow = F,  xanchor='right', yanchor='auto', xshift=0, yshift=0,font = list(size = 7)))

    ```
